### 2012-09-14 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to direct method section

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{Eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations for the \textit{Eigenvectors} of
$\ve{A}$. A quantity of particular interest that is computatable from
the eigenvalues of a matrix is the \textit{spectral radius},
$\rho(\ve{A})$ defined by Saad \citep{saad_2003} as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}
In addition, for problems that have a large scale over which the
independent variables may exist (e.g. a problem with events on
timescales ranging from nanoseconds to hours), a good measure of this
range is supplied by the \textit{stiffness ratio}:
\begin{equation}
  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
\end{equation}
Those problems that have a wide range of scales in their independent
variables, which will then be reflected in the operator, will then
have a large stiffness ratio. We will define such problems with large
stiffness ratios as \textit{stiff}.

@@

We can also compute the norm of a matrix by inferring from the norm of
the vector on which it is operating. Per LeVeque, we search for a
constant that is equivalent to $||\ve{A}||$:
\begin{equation}
  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
  \label{eq:matrix_norm_inequality}
\end{equation}
where the minimum value of $C$ that satisfies
Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
that is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
simple terms for common norms including the infinity norm:
\begin{equation}
  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
  \label{eq:matrix_infinity_norm}
\end{equation}
and the 2-norm:
\begin{equation}
  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
  \label{eq:matrix_2_norm}
\end{equation}
where $\rho$ is the spectral radius as defined in
Eq~(\ref{eq:spectral_radius}).

@@

\section{Iterative Methods for Solving Sparse Linear Systems}
\label{sec:linear_methods}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
elements \citep{zienkiewicz_1977} ultimately generate matrix problems
as in Eq~(\ref{eq:linear_problem}).  In addition, these matrices are
typically sparse, meaning that the vast majority of their constituent
elements are zero. This sparsity is due to the fact that the influence
of a particular grid element only expands as far as a few of its
nearest neighbors depending on the order of discretization used, and
therefore coupling among variables in a particular discrete equation
in the system leads to a few non-zero entries. Because of the natural
occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods to provide
background and then give detail on a stochastic class of methods.

\subsection{Stationary Methods}
\label{subsec:stationary_methods}
Stationary methods arise from splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{A} = \ve{M} - \ve{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
particular method chosen. Using this split definition of the operator
we can then write:
\begin{equation}
  \ve{M}\ve{x} + \ve{N}\ve{x} = \ve{b}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
    \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index}. In
general, we will define methods in the form of
Eq~(\ref{eq:linear_iterative_method}) as \textit{stationary
  methods}. Given this, we can then generate a few statements
regarding the convergence of such stationary methods. Defining
$\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution error at the
$k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
from Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form
of the linear problem:
\begin{equation}
  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. We can then
place restrictions on the iteration matrix by using the tools
developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
is diagonalizable (although we may generalize this with the Jordan
canonical form of $H$ \citep{saad_2003}, we then have:
\begin{equation}
  \ve{e}^{k} =
  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
$\ve{H}$. Computing the 2-norm of the above form then gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\ve{R})=0$ and the error bound reduces to:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now resrtict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
for a stationary method to converge such that $k$ applications of the
iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}).

\subsection{Projection Methods}
\label{subsec:projection_methods}
Among the most common iterative methods used in scientific computing
today for sparse systems are a broad class known as \textit{projection
  methods}. These methods not only provide access to more powerful
means of reaching a solution, but also a powerful means of
encapsulating the majority of common iterative methods including the
stationary methods just discussed in a common mathematical
framework. All projection methods are built around a core structure
where the solution to Eq~(\ref{eq:linear_problem}) is extracted from a
\textit{search subspace} $\mathcal{K}$ and bound by a
\textit{constraint subspace} $\mathcal{L}$ that will vary in
definition depending on the iterative method selected. We build the
approximate solution $\tilde{\ve{x}}$ by starting with an initial
guess $\ve{x}_0$ and extracting a correction $\boldsymbol{\delta}$
from $\mathcal{K}$ such that:
\begin{equation}
  \tilde{\ve{x}} = \ve{x}_0 +
  \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}\:.
  \label{eq:linear_projection_step}
\end{equation}
We bound this correction by asserting that the new residual,
$\tilde{\ve{r}}$, be orthogonal to $\mathcal{L}$:
\begin{equation}
  \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
  \mathcal{L}\:.
  \label{eq:linear_projection_constraint}
\end{equation}

@@

From an iteration standpoint, as we choose $\boldsymbol{\delta}$ from
$\mathcal{K}$ and constrain it with $\mathcal{L}$, each iteration
performs a projection that systematically anihilates the components of
the residual that exists in $\mathcal{L}$. This then means that if our
convergence criteria for an iterative method is bound the residual of
the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
projection step gaurantees us that the norm of the new residual will
never be worse than that of the previous step and will typically move
us towards convergence. Depending on the qualities of the system in
Eq~(\ref{eq:linear_problem}), the selection of the subspaces
$\mathcal{K}$ and $\mathcal{L}$ can serve to both gaurantee
convergence and optimize the rate at which the residual is decreased.

@@

Knowing the projections onto the search and constraint subspaces, the
GMRES method may be formulated. We note here several properties of
this formulation and how they may facilitate or hinder the solution of
large-scale, sparse linear problems, also noting that these properties
are common among many Krylov methods. First, from a memory perspective
GMRES is efficient from the standpoint that the operator $\ve{A}$ need
not be explicitly stored. Rather, only the ability to compute the
action of that operator on a vector of valid size is
required. However, these savings in memory are balanced by the fact
that the long recurrence relations used in the Arnoldi procedure
require all vectors that span the Krylov space to be stored. If the
size of these vectors becomes prohibitive, the Arnoldi procedure can
be restarted at the cost of losing information in the
orthogonalization process, creating the potentitial to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_2010}) that may
not be amenable to its construction.

\subsection{Stochastic Methods}
\label{subsec:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable 1950 manuscript by Forsythe and Leibler
\citep{forsythe_1950}. In their outline, Forsythe and Liebler in fact
credit the creation of this technique to J. Von Neumann and S.M. Ulam
some years earlier than its publication. Hammersley and Handscomb's
1964 monograph \citep{hammersley_1964} presents additional detail on
this topic using a collection of references from the 1950's and early
1960's.

We begin our discussion of stochastic methods using these texts by
seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
operator $\ve{A}$, we can define the following operator:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:.
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this inverse sum to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation will therefore lead to an
approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering its $i^{th}$
component:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states
$i_{m-1}$ and $i_{m}$ in $\ve{H}$. We can generate these transitions
through Monte Carlo random walks by assigning them both a probability
and weight. Knowing this, we then form the \textit{Neumann-Ulam
  decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Schur product operation, $\ve{P}$ denotes
the transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}).

\subsubsection{Direct Method}
\label{subsubsec:direct_mc}
In the context of matrix inversion, a direct method resembles an
adjoint Monte Carlo method in the reactor physics community where the
solution state is sampled and the source terms that contribute to it
assembled. To achieve this, we build the direct Neumann-Ulam
decomposition by first choosing a probability matrix that is a column
scaling of $\ve{H}$ such that its components are:
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
  \label{eq:direct_probability}
\end{equation}
From this, we then see that the probability of transitioning from a
state $i$ to a state $j$ is implicitly linked to the original operator
$\ve{A}$ in that those terms with large values, and therefore make the
greatest contribution to the numerical solution, will be sampled with
a higher probability than smaller terms. In addition, the column
scaling provides a normalization over the state to which we are
transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
the probabilities over the columns of the matrix. The components of
the weight matrix are then defined by
Eq~(\ref{eq:neumann_ulam_decomposition}) as:
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:direct_weight}
\end{equation}
It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
$\ve{P}$, and $\ve{W}$ must be sparse as well by definition.

Using these matrices, we can then form the expectation value of the
solution.

>>>>

In a discussion of methods for solving linear systems, several
mathematical tools are useful in characterizing the qualities of the
linear system. Among the most useful are the \textit{Eigenvalues} of
the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
problem:
\begin{equation}
  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
  \label{eq:eigenvalue_problem}
\end{equation}
By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
\begin{equation}
  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
  \label{eq:eigenvalue_problem_2}
\end{equation}
and demanding that non-trivial solutions for $\ve{x}$ exist, it is
then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
determinant yields a characteristic polynomial in terms of $\lambda$
with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
component of $\sigma(\ve{A})$ can then be used to solve
Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
of $\ve{A}$. A quantity of particular interest that is computatable
from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_2003} as:
\begin{equation}
  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
  \label{eq:spectral_radius}
\end{equation}
In addition, for problems that have a large scale over which the
independent variables may exist (e.g. a problem with events on
timescales ranging from nanoseconds to hours), a good measure of this
range is supplied by the \textit{stiffness ratio}:
\begin{equation}
  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
\end{equation}
Those problems that have a wide range of scales in their independent
variables, which will then be reflected in the operator, will then
have a large stiffness ratio. We will define such problems with large
stiffness ratios as \textit{stiff}.

@@

We can also compute the norm of a matrix by inferring from the norm of
the vector on which it is operating. Per LeVeque, we search for a
constant that is equivalent to $||\ve{A}||$:
\begin{equation}
  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
  \label{eq:matrix_norm_inequality}
\end{equation}
where the minimum value of $C$ that satisfies
Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
simple terms for common norms including the infinity norm:
\begin{equation}
  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
  \label{eq:matrix_infinity_norm}
\end{equation}
and the 2-norm:
\begin{equation}
  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
  \label{eq:matrix_2_norm}
\end{equation}
where $\rho$ is the spectral radius as defined in
Eq~(\ref{eq:spectral_radius}).

@@

\section{Iterative Methods for Solving Sparse Linear Systems}
\label{sec:linear_methods}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
elements \citep{zienkiewicz_1977} ultimately generates matrix problems
as in Eq~(\ref{eq:linear_problem}).  In addition, these matrices are
typically sparse, meaning that the vast majority of their constituent
elements are zero. This sparsity is due to the fact that the influence
of a particular grid element only expands as far as a few of its
nearest neighbors depending on the order of discretization used and
therefore coupling among variables in a particular discrete equation
in the system leads to a few non-zero entries. Because of the natural
occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide background and then give detail on a
stochastic class of methods.

\subsection{Stationary Methods}
\label{subsec:stationary_methods}
Stationary methods arise from splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{A} = \ve{M} - \ve{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
particular method chosen. Using this split definition of the operator
we can then write:
\begin{equation}
  \ve{M}\ve{x} + \ve{N}\ve{x} = \ve{b}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
    \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index}. In
general, we will define methods in the form of
Eq~(\ref{eq:linear_iterative_method}) as \textit{stationary
  methods}. Given this, we can then generate a few statements
regarding the convergence of such stationary methods. Defining
$\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution error at the
$k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
from Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form
of the linear problem:
\begin{equation}
  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. We can then
place restrictions on the iteration matrix by using the tools
developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
is diagonalizable\footnote{We may generalize this to
  non-diagonalizable matrices with the Jordan canonical form of
  $\ve{H}$.} \citep{saad_2003}, we then have:
\begin{equation}
  \ve{e}^{k} =
  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
$\ve{H}$. Computing the 2-norm of the above form then gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\ve{R})=1$ and the error bound reduces to:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now resrtict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
for a stationary method to converge such that $k$ applications of the
iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}).

\subsection{Projection Methods}
\label{subsec:projection_methods}
Among the most common iterative methods used in scientific computing
today for sparse systems are of a broad class known as
\textit{projection methods}. These methods not only provide access to
more powerful means of reaching a solution, but also a powerful means
of encapsulating the majority of common iterative methods including
the stationary methods just discussed in a common mathematical
framework. All projection methods are built around a core structure
where the solution to Eq~(\ref{eq:linear_problem}) is extracted from a
\textit{search subspace} $\mathcal{K}$ and bound by a
\textit{constraint subspace} $\mathcal{L}$ that will vary in
definition depending on the iterative method selected. We build the
approximate solution $\tilde{\ve{x}}$ by starting with an initial
guess $\ve{x}_0$ and extracting a correction $\boldsymbol{\delta}$
from $\mathcal{K}$ such that:
\begin{equation}
  \tilde{\ve{x}} = \ve{x}_0 +
  \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}\:.
  \label{eq:linear_projection_step}
\end{equation}
We bound this correction by asserting that the new residual,
$\tilde{\ve{r}}$, be orthogonal to $\mathcal{L}$:
\begin{equation}
  \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
  \mathcal{L}\:.
  \label{eq:linear_projection_constraint}
\end{equation}

@@

From an iteration standpoint, as we choose $\boldsymbol{\delta}$ from
$\mathcal{K}$ and constrain it with $\mathcal{L}$, each iteration
performs a projection that systematically anihilates the components of
the residual that exists in $\mathcal{L}$. This then means that if our
convergence criteria for an iterative method is bound to the residual
of the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
projection step gaurantees us that the norm of the new residual will
never be worse than that of the previous step and will typically move
us towards convergence. Depending on the qualities of the system in
Eq~(\ref{eq:linear_problem}), the selection of the subspaces
$\mathcal{K}$ and $\mathcal{L}$ can serve to both gaurantee
convergence and optimize the rate at which the residual is decreased.

@@

Knowing the projections onto the search and constraint subspaces, the
GMRES method may be formulated. We note here several properties of
this formulation and how they may facilitate or hinder the solution of
large-scale, sparse linear problems, also noting that these properties
are common among many Krylov methods. First, from a memory perspective
GMRES is efficient in that the operator $\ve{A}$ need not be
explicitly stored. Rather, only the ability to compute the action of
that operator on a vector of valid size is required. However, these
savings in memory are balanced by the fact that the long recurrence
relations used in the Arnoldi procedure require all vectors that span
the Krylov space to be stored. If the size of these vectors becomes
prohibitive, the Arnoldi procedure can be restarted at the cost of
losing information in the orthogonalization process, creating the
potentitial to generate new search directions that are not orthogonal
to all previous search directions (and therefore less than
optimal). From an implementation perspective, because the operator is
not required to be formed, GMRES is significantly more flexible in its
usage in that there are many instances where various processes serve
to provide the action of that operator (e.g. radiation transport
sweeps \citep{evans_2010}) that may not be amenable to its
construction.

\subsection{Stochastic Methods}
\label{subsec:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_1950}. In their outline, Forsythe and Liebler in fact
credit the creation of this technique to J. Von Neumann and S.M. Ulam
some years earlier than its publication. Hammersley and Handscomb's
1964 monograph \citep{hammersley_1964} presents additional detail on
this topic using a collection of references from the 1950's and early
1960's.

We begin our discussion of stochastic methods using these texts by
seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
operator $\ve{A}$, we can split it in the same manner as
Eq~(\ref{eq:linear_split_equation2}) to define the following operator:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:.
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this inverse sum to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation will therefore lead to an
approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering its $i^{th}$
component:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is a particular sequence permutation. We can
generate these sequences of transitions through Monte Carlo random
walks by assigning them both a probability and weight. As a
reinterpretation of the iteration matrix, we then form the
\textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The Hadamard
  product $\ve{A} = \ve{B} \circ \ve{C}$ is defined element-wise as
  $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the transition
probability matrix, and $\ve{W}$ denotes the transition weight
matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be dependent on
whether we choose a direct or adjoint Monte Carlo sequence to estimate
the state transitions in Eq~(\ref{eq:expanded_neumann_solution}).

\subsubsection{Direct Method}
\label{subsubsec:direct_mc}
In the context of matrix inversion, a direct method resembles an
adjoint Monte Carlo method in the reactor physics community where the
solution state is sampled and the source terms that contribute to it
are assembled. To achieve this, we build the direct method
Neumann-Ulam decomposition by first choosing a probability matrix that
is a column scaling of $\ve{H}$ such that its components are:
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
  \label{eq:direct_probability}
\end{equation}
From this, we then see that the probability of transitioning from a
state $i$ to a state $j$ is implicitly linked to the original operator
$\ve{A}$ in that those terms with large values, and therefore those
that make the greatest contribution to the numerical solution, will be
sampled with a higher probability than smaller terms. In addition, the
column scaling provides a normalization over the state to which we are
transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
the probabilities over the columns of the matrix. The components of
the weight matrix are then defined by
Eq~(\ref{eq:neumann_ulam_decomposition}) as:
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:direct_weight}
\end{equation}
It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
$\ve{P}$, and $\ve{W}$ must be sparse as well by
definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
the non-zero elements of $\ve{H}$ as those components that are zero
will not participate in the random walk. Furthermore, it prevents an
infinite weight from being generated in Eq~(\ref{eq:direct_weight}).

Using these matrices, we can then form the expectation value of the
direct solution. For a given random walk permutation $\nu$ with $k$
events, we define the weight of that permutation to be:
\begin{equation}
  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
such that the weight of each transition event contributes to the
total. The contribution to the solution from a particular random walk
permutation is then:
\begin{equation}
  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
  \label{eq:direct_permutation_contribution}
\end{equation}
where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
are tallying defines state $i$.  We can interpret this precisely as
before in that during the random walk we collect the source in the
states that are visited and apply them to the solution tally. We then
define the probability that a particular random walk permuation of $k$
events will occur:
\begin{equation}
  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
  \label{eq:direct_permutation_probability}
\end{equation}
Finally, we define the expectation value of $X$ to be the collection
of all random walk permutations and their probabilities:
\begin{equation}
  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{p\nu}\:,
  \label{eq:direct_expectation_value}
\end{equation}
which, if expanded, directly recovers the exact solution:
\begin{multline}
  E\{X(i_0 = i)\}\\
  =\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ = x_i\:,
  \label{eq:direct_expectation_expansion}
\end{multline}
therefore providing an unbiased Monte Carlo estimator.
